---
output:
  pdf_document:
    includes:
      in_header: subfiles/import.tex
      before_body: subfiles/before_body.tex

bibliography: subfiles/references.bib
nocite: '@*'
---

```{r, include=FALSE}
# Copyright (c) 2022 Vincent Talen.
# Licensed under GPLv3. See LICENSE file.
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(fig.align = "center")
```

\newpage
<!-- (Introduction) -->
# Introduction
<!-- Introduction of the research and introduction research questions -->

## Goal
<!-- - Describe Goal (not the educational goal but the research goal) -->
<!-- - Describe how you reach the goal (e.g. make model and figures, use different setting) -->
<!-- - formulate hypothesis -->

## Theory
<!-- - Describe biological model -->
<!-- - Picture of the biological model -->

<!-- Give an explanation of the model with citations of source `\cite{Soertaert10}` and formula explanation. -->

<!-- Describe each element and the transformations -->


\newpage
<!-- (Methods) -->
# Methods
## The software model
<!-- - Describe the software tools used, as well as the libraries -->
<!-- - Describe the software implementation -->

## Model configuration
<!-- Explain chosen initial state, parameter values and time sequence. Use tables with values. -->

RMR was calculated as the slope of oxygen depletion over a ca 35 min timeframe, corrected for the mean linear trend over time of oxygen concentration in controls as follows: RMR = (ci – cf)/(ti – tf), where ci and cf are initial and final oxygen concentrations (μmol O2/L) in the wells and ti and tf are initial and final times points (day-1) of the respiration experiment. Values were converted into C release rate (μg C/day) assuming a respiratory coefficient of 0.78 (as estimated for Gammarus pulex, Wright & Wright, 1979).

### Statistical Analysis Equations
The following equations were used to express the mass (M in mg) and temperature (T in Kelvin) dependence of individual RMR and IR:

\mtestandard
\mtequadratic

The standard MTE formulation (\ref{eq:mte_standard}) is simply a particular case of the quadratic formulation (\ref{eq:mte_quadratic}) where $q = 0$ and 
the equation is reduced to the MTE model where p can thus be interpreted as the activation energy.

Energetic efficiency was also calculated as follows: $E = (IR/RMR) * A_T$, where the ratio of IR to RMR is the ingestion to metabolism efficiency and $A_T$ is the assimilation efficiency at temperature $T$. The temperature ($T$ in Kelvin) dependence of assimilation efficiency was expressed, using empirical equations and values for detritivores Assimilation efficiency was following a logistic equation with the MTE equation both at the numerator and the denominator
With the following formulation, assimilation efficiency is confined between 0 and 1 (no or complete assimilation):

\assimilationefficieny

### Consumer-Resource Dynamics Model
Below are the ordinary differential equations describing temporal change in leaf litter standing stocks ($L$) and Gammarus population biomass ($G$) (Equation \ref{eq:ode_leaf_litter} and \ref{eq:ode_gammarus_population}).

\leaflitter
\gammaruspopulation


\newpage
<!-- (Results) -->
# Results
<!-- Introduction of results, how does it answer your research questions.   -->

<!-- - Describe what can be seen in such way that it leads to an answer to your research questions -->
<!-- - Give your figures a number and a descriptive title. -->
<!-- - Provide correct axis labels (unit and quantity), legend and caption. -->
<!-- - Always refer to and discuss your figures and tables in the text - they never stand alone. -->


\newpage
<!-- (Discussion and Conclusion) -->
# Discussion and Conclusion
## Discussion
<!-- - Compare your results with what is expecting from the literature and discuss differences with them. -->
<!-- - Discuss striking and surprising results. -->
<!-- - Discuss weaknesses in your research and how they could be addressed. -->
- Basically unreproducible in general, long pieces of code that were repeated over and over that could not be understood. No easy way to reproduce other than to immediately copy and paste multiple hundreds lines of code where only few values would be changed.
- Formulas for metabolic and ingestion rates were heavily rewritten from the base formula. They were unrecognizable so they were cleaned up to be understandable and resemble the actual formula more. It should also be noted that the position where the mean in the quadratic portion of the exponent is actually different from what would be done following the formula.

## General conclusion and perspective
<!-- Discuss what your goal was, what the end result is and how you could continue working from here. -->


<!-- (After-Body) -->
```{r after body, child = 'subfiles/after_body.Rmd'}
```
