---
title: "Assignment week 1"
subtitle: "mRNA Dynamic"
output:
  pdf_document

header-includes:
   - \pagenumbering{gobble}
   - \usepackage{longtable}
   - \usepackage{hyperref}
---

```{r setup, include=FALSE}
# Set code chunks visibility in pdf output to true
knitr::opts_chunk$set(echo = TRUE)

# Create vector with all packages that will be used
packages <- c("deSolve")
# Load each package in the vector with lapply
invisible(lapply(packages, library, character.only = TRUE))
# Drop the packages variable from memory since it will not be used again
remove(packages)
```


<!-- (Front page) -->
\vspace{60pt}
\begin{center}
  \includegraphics[width=90mm]{../data_weekly_assignments/hanze.png}
\end{center}
\vspace{45pt}

\hfill \textbf{Student}: Vincent Talen  

\hfill \textbf{Student number}: 389015  

\hfill \textbf{Class}: BFV2

\hfill \textbf{Study}: Bio-Informatics

\hfill \textbf{Institute}: Institute for Life Science & Technology

\hfill \textbf{Teacher}: Tsjerk Wassenaar

\hfill \textbf{Date}: `r Sys.Date()`


<!-- (Page 01 - Table of contents) -->
\newpage
\setcounter{secnumdepth}{2}
\tableofcontents
\pagenumbering{arabic}


<!-- (Assignment 1) -->
\newpage
# Assignment 1
## 1.1 {-}
There are two parameters that need to be programmed, namely;
  - r (rate of decay of existing transcripts in percentage)
  - m (amount of new transcripts per second)

## 1.2 {-}
Differential equations first came into existence with the invention of calculus by Newton and Leibniz \cite{Wikipedia1}.
This specific model equation was made to calculate differences in mRNA transcripts present that are only influenced by a rate of decay and the amount of transcripts that get created each second.

## 1.3 {-}
$$ \frac{\Delta R}{\Delta t} = -r * R + m $$
\begin{center}
  \includegraphics[width=90mm]{../data_weekly_assignments/biological_model_week1.png}
\end{center}

## 1.4 {-}
Below is the model function that can be used for the `ode` function from the `deSolve` package, it returns the delta R (the change in number of mRNA transcripts present).
The delta R is returned and not the actual R number is that the `ode` function needs the model to return it in a specific way and itself calculates the resulting R values. The change for a specific time frame needs to be returned.

```{r create model function}
# Create model function
model <- function(time, cur.state, parameters) {
  # Unpack the cur.state (R) and the parameters (r & m) for instant access
  with(as.list(c(cur.state, parameters)),{
    # Calculate change in R and return it
    delta.R <- -r * R + m
    return(list(delta.R))
    })
}
```


<!-- (Assignment 2) -->
# Assignment 2
```{r calculate and plot mRNA dynamics of 3 situations with different parameters}
# Initial value and time frame
initial.values <- c(R = 100)
times <- seq(0, 15, by = 0.01)

# Increasing R
incr.params <- c(r = 0.5, m = 60)
incr.out <- ode(y = initial.values, times = times, func = model, 
                parms = incr.params, method = "euler")

# Steady state
steady.params <- c(r = 0.5, m = 50)
steady.out <- ode(y = initial.values, times = times, func = model, 
                  parms = steady.params, method = "euler")

# Decreasing R
decr.params <- c(r = 0.5, m = 40)
decr.out <- ode(y = initial.values, times = times, func = model, 
                parms = decr.params, method = "euler")

# Create plot with steady state and add increasing- and decreasing states after
plot(steady.out, type = 'l', ylim = c(75,125), col='green',
     main = 'Amount of mRNA transcripts over time',
     xlab = "Timepoints", ylab = "number of mRNA transcripts")
lines(incr.out, col='red')
lines(decr.out, col='blue')

# Add a legend to the plot
legend(9, 117, legend = c('steady state', 'increasing over time', 'decreasing over time'), 
       fill = c('red', 'green', 'blue'))
```

<!-- (References) -->
\newpage
\begin{thebibliography}{9}

\bibitem{Wikipedia1}
Wikipedia contributors: \textit{Differential equation}, Wikipedia, The Free Encyclopedia, \url{https://en.wikipedia.org/w/index.php?title=Differential_equation&oldid=1087007437} (accessed May 10, 2022).

\end{thebibliography}
