---
title: "Assignment week 3"
subtitle: "Detailed analysis of the glucocorticoid receptor dynamica model"
output:
  pdf_document

header-includes:
   - \pagenumbering{gobble}
   - \usepackage{longtable}
   - \usepackage{hyperref}
---

```{r setup, include=FALSE}
# Set code chunks visibility in pdf output to true
knitr::opts_chunk$set(echo = TRUE)

# Create vector with all packages that will be used
packages <- c("deSolve", "ggplot2", "reshape2", "gridExtra")
# Load each package in the vector with lapply
invisible(lapply(packages, library, character.only = TRUE))
# Drop the packages variable from memory since it will not be used again
remove(packages)
```


<!-- (Front page) -->
\vspace{60pt}
\begin{center}
  \includegraphics[width=90mm]{../hanze.png}
\end{center}
\vspace{45pt}

\hfill \textbf{Student}: Vincent Talen  

\hfill \textbf{Student number}: 389015  

\hfill \textbf{Class}: BFV2

\hfill \textbf{Study}: Bio-Informatics

\hfill \textbf{Institute}: Institute for Life Science & Technology

\hfill \textbf{Teacher}: Tsjerk Wassenaar

\hfill \textbf{Date}: `r Sys.Date()`


<!-- (Table of contents) -->
\newpage
\setcounter{secnumdepth}{2}
\tableofcontents
\pagenumbering{arabic}


<!-- (Assessing Model Validity) -->
\newpage
# Assignment 1: Assessing Model Validity
To assess the validity of the model, the experimental data will be compared to the simulation data from previous week.
The comparison will be done by plotting the experimental data and the simulation data in the same graph.

## Loading in the data
```{r load data from file}
# Load in data from file
data <- read.csv("MPL.csv", na.strings = "NA")
# Rename 'mRNA' and 'Free_receptor' columns to use the same name scheme
names(data)[4:5] <- c("Rm", "R")
```

The actual experiment data contains multiple values per time, if a plot is made with those raw values the plot below will result.
```{r show raw values plot, warning=FALSE, out.width = "66%", fig.align = "center", fig.cap = "Demonstration of plotting raw data"}
# Create plot with raw values to demonstrate the need of using medians
ggplot(data, mapping = aes(x = time, y = R, width = 2)) + geom_line() + geom_point()
```
As can be seen the line goes through each single point and thus is the line not that useful. 
So to be able to get a meaningful plot the medians are used to create a useful line of the medians that can then be used to compare the values.

```{r create data frame with medians}
# Create new data frame containing the medians per time
data.medians <- aggregate(data[,c("MPL_conc","Rm","R")],
                          list(data$dose,data$time), median, na.rm=T)
names(data.medians)[1:2] <- c("dose","time")
```

\newpage
## Implementing GRD model with given values

| Parameter | Value | Unit               | Explanation
|----       |---    |------              |--------------------
| Rm        | 4.74  | $fmol/g\ liver$    | concentration of receptor mRNA
| R         | 267   | $fmol/mg\ protein$ | concentration of free receptor in cytosol
| DR        | 0     | $fmol/mg\ protein$ | concentration of receptor complex in cytosol
| DR_N      | 0     | $fmol/mg\ protein$ | concentration of receptor complex in nucleus
Table: Initial Values for MPL

| Parameter | Value   | Unit               | Explanation
|----       |:---:    |:------:            |--------------------
| k.s_Rm    | 2.9     | $fmol/g\ liver/h$  | zero-order rate constant of receptor mRNA synthesis
| k.d_Rm    | 0.612   | -                  | first-order rate constant receptor mRNA degradation
| IC.50_Rm  | 26.2    | $fmol/mg\ protein$ | concentration of DR_N where receptor mRNA synthesis drops to 50% of base value
| k.on      | 0.00329 | $L/nmol/h$         | second-order rate constant of receptor complex formation
| k.T       | 0.63    | $1/h$              | first-order rate constant of translocation of receptor complex to nucleus
| k.re      | 0.57    | $1/h$              | first-order rate constant of receptor 'recovery' from nucleus to cytosol
| R.f       | 0.49    | -                  | fraction of receptor being recycled from complexes
| k.s_R     | 3.22    | -                  | first-order rate constant of receptor synthesis
| k.d_R     | 0.0572  | $1/h$              | first-order rate constant of receptor degradation
| D         | -       | $nmol/L$           | plasma concentration of corticosteroid
Table: Parameter Values for MPL

```{r use previously created model to generate simulation data}
# Create function of GRD model
model <- function(time, cur.state, parameters) {
  # Unpack the current state and the parameters for instant access
  with(as.list(c(cur.state, parameters)), {
    # Calculate delta for each equation and return them in a list
    delta.Rm <- k.s_Rm * ( 1 - DR_N / ( IC.50_Rm + DR_N )) - k.d_Rm * Rm
    delta.R <- k.s_R * Rm + R.f * k.re * DR_N - k.on * D * R - k.d_R * R
    delta.DR <- k.on * D * R - k.T * DR
    delta.DR_N <- k.T * DR - k.re * DR_N
    return(list(c(delta.Rm, delta.R, delta.DR, delta.DR_N)))
    })
}

# Initial values and time frame
initial.values <- c(Rm = 4.74, R = 267, DR = 0, DR_N = 0)
times <- seq(0, 168, by = 1)

# Create function that converts concentration of MPL from ng/mL to nmol/L
get.D <- function(ng.ml.concentration) { return(ng.ml.concentration * 1000 / 374.471) }
# Define parameters determined for methylprednisolone (MPL)
parameters <- c(k.s_Rm = 2.9, k.d_Rm = 0.612, IC.50_Rm = 26.2, k.on = 0.00329, k.T = 0.63, 
                k.re = 0.57, R.f = 0.49, k.s_R = 3.22, k.d_R = 0.0572, D = get.D(20))
```

## Creating plots to compare experiment data with simulation data
```{r get the median mpl concentrations for the doses}
# Get all unique doses present in the data except 0.0
all.doses <- unique(data$dose)
all.doses <- all.doses[!all.doses %in% (0.0)]

# Create function that calculates median MPL concentration for a dose
get.median.mpl_conc <- function(i) { median(data$MPL_conc[data$dose==i], na.rm=T) }

# Create data frame with mpl concentration for each dose
median.mpl_conc <- data.frame("dose" = all.doses, 
                              "mpl_conc" = unlist(lapply(all.doses, get.median.mpl_conc)))
```

```{r}
get_legend<-function(myggplot){
  tmp <- ggplot_gtable(ggplot_build(myggplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)
}
```

```{r create plots, warning=FALSE, fig.align = "center", fig.height = 2.75}
for (current.dose in all.doses) {
  # Change to correct D parameter for current dose
  parameters$D <- get.D(median.mpl_conc$mpl_conc[median.mpl_conc$dose == current.dose])
  
  # Get subset of experiment data for current dose
  cur.data <- subset(data, dose == 0.0 | dose == current.dose)
  # Get subset of experiment median data for current dose
  cur.data.medians <- subset(data.medians, dose == 0.0 | dose == current.dose)
  # Perform ODE function using the model with the given values for MPL
  cur.data.simulated <- as.data.frame(ode(times = times, y = initial.values, 
                                       parms = parameters, func = model, method = "euler"))
  
  # Create plot for receptor mRNA
  plot.Rm <- ggplot(cur.data.simulated, aes(x = time, y = Rm)) + 
    geom_line(aes(color = "Simulation Data")) +
    geom_line(aes(color = "Experiment Median"), data = cur.data.medians) + 
    geom_point(aes(color = "Experiment Data"), data = cur.data, shape = 1) + 
    ggtitle("Receptor mRNA") + ylab("Rm (fmol/g liver)") + xlab("time (h)") + 
    theme(legend.position = "bottom", plot.margin = margin(0.25, 0.25, 0.25, 0.25, "cm")) + 
    scale_color_manual("", 
                       limits=c("Simulation Data", "Experiment Median", "Experiment Data"), 
                       values = c("black","red","black")) +
    guides(color = guide_legend(override.aes = list(linetype = c(1, 1, NA), shape = c(NA, NA, 1))))
  
  # Create plot for free receptor concentration
  plot.R <- ggplot(cur.data.simulated, aes(x = time, y = R)) +
    geom_line(aes(color = "Simulation Data")) +
    geom_line(aes(color = "Experiment Median"), data = cur.data.medians) +
    geom_point(aes(color = "Experiment Data"), data = cur.data, shape = 1) + 
    ggtitle("Free receptor concentration") + ylab("R (fmol/mg protein)") + xlab("time (h)") +
    theme(legend.position = "bottom", plot.margin = margin(0.25, 0.25, 0.25, 0.25, "cm")) + 
    scale_color_manual("", limits=c("Simulation Data", "Experiment Median", "Experiment Data"), 
                       values = c("black","red","black")) +
    guides(color = guide_legend(override.aes = list(linetype = c(1, 1, NA), shape = c(NA, NA, 1))))
  
  # Extract legend
  legend <- get_legend(plot.R)
  # Remove the legends from the original plots
  plot.R <- plot.R + theme(legend.position="none")
  plot.Rm <- plot.Rm + theme(legend.position="none")
  
  grid.arrange(plot.Rm, plot.R, legend, ncol = 2, nrow = 2,
             layout_matrix = rbind(c(1,2), c(3,3)),
             widths = c(2.7, 2.7), heights = c(2.5, 0.2),
             top = sprintf("Graphs for dose = %s (mg drug/kg rat/h)", current.dose))
}
```

## Results of plot comparison
[2] How do the results of the simulations depend on the dose and concentration of the drug?  
Compare the model variables mRNA, R with the experimental data
    
[3] Are the results of the model in line with experimental data? If not, what could be the reason? Think of at least one explanation.  
Try to test it with simulations (you will get bonus points for that, your explanation does not need to be correct, but should be logical).


<!-- (References) -->
\newpage
\begin{thebibliography}{9}

\bibitem{Wikipedia1}
Barnes, P.J. (2011), \textit{Glucocorticosteroids: current and future directions}, British Journal of Pharmacology, 163: 29-43, \url{https://doi.org/10.1111/j.1476-5381.2010.01199.x} (accessed May 11, 2022).

\end{thebibliography}
